---
vignette: >
  %\VignetteIndexEntry{Subset and aggregate multidimensional data with data.cube}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Subset and aggregate multidimensional data with `data.cube`
*Jan Gorecki, 2016-05-04*

R *data.cube* class defined in [data.cube](https://gitlab.com/jangorecki/data.cube) package.  

## characteristic of `[.data.cube`

### what's the same in `[.array`

- subsetting dimensions by keys
- handling *drop* argument
- returns same class object, so it is *chain-able*

### what's more than `[.array`

- can subset on hierarchy attributes of dimensions
- scale with dimensions and their cardinalities
- store multiple measures, predefined formatting
- high performance thanks to `data.table`
- can aggregate measures by dropping chosen dimensions

### what's different to `[.array`

- array can be subsetted by dimension keys but also by integer position
- in data.cube subset directly by integer position is not possible, but you can have integer surrogate keys which can refer to any custom integer positions
- dimension levels are always ordered

## start session

```{r run}
# install.packages("data.cube", repos = paste0("https://", c("jangorecki.gitlab.io/data.cube","Rdatatable.github.io/data.table","cran.rstudio.com")))
library(data.table)
library(data.cube)
```

## Tiny example

```{r tiny}
set.seed(1)
# array
ar = array(rnorm(8,10,5), rep(2,3), 
           dimnames = list(color = c("green","red"), 
                           year = c("2014","2015"), 
                           country = c("IN","UK"))) # sorted
# cube normalized to star schema just on natural keys
dc = as.data.cube(ar)
# slice
ar["green", "2015",]
dc["green", "2015"]
format(dc["green", "2015"])
# dice
ar[c("green","red"), c("2014","2015"),]
dc[c("green","red"), c("2014","2015")]
format(dc[c("green","red"), c("2014","2015")])

# exact tabular representation of array is just a formatting on the cube
ar["green", c("2014","2015"),]
format(dc["green", c("2014","2015")], 
       dcast = TRUE, 
       formula = year ~ country)

ar[,"2015", c("UK","IN")]
format(dc[,"2015", c("UK","IN")], 
       dcast = TRUE, 
       formula = color ~ country) # sorted dimensions levels
```

## Hierarchies example

```{r hierarchies}
# as.data.cube.list - investigate X to see structure
X = populate_star(N=1e5)
lapply(X[1:2], sapply, ncol)
lapply(X[1:2], sapply, nrow)
str(X[3L], max.level=3L) # hierarchy defined as list of levels and attributes on each level
dc = as.data.cube(X)
print(dc)

# slice
dc["Mazda RX4"]
dc["Mazda RX4",,"BTC"]
# dice
dc[,, c("CNY","BTC"), c("GA","IA","AD")]

# custom order of arguments matching by dimension name
dc[geography = c("GA","IA","AD"), currency = c("CNY","BTC")]

# check dimensions
names(dc$dimensions)
# foreign keys and measures
names(dc)

# use dimensions hierarchy attributes for slice and dice, mix filters from various levels in hierarchy
dc["Mazda RX4",, .(curr_type = "crypto"),, .(time_year = 2014L, time_quarter_name = c("Q1","Q2"))]
# same as above but more verbose
dc[product = "Mazda RX4",
   customer = .(),
   currency = .(curr_type = "crypto"),
   geography = .(),
   time = .(time_year = 2014L, time_quarter_name = c("Q1","Q2"))]

# data.cube `[` operator returns another cube so queries can be chained
dc[,,, .(geog_region_name = "North Central")
   ][,,, .(geog_abb = c("IA","NV","MO")), .(time_year = 2014L)
     ]
```

## Aggregates while subsetting

```{r aggregate}
# aggregate by droppping dimension with `-` symbol, group by customer and currency
dc[product = `-`,
   customer = .(),
   currency = .(curr_type="crypto"),
   geography = `-`,
   time = `-`(time_year = 2013:2014)]
```

## scalability

```{r scalability}
# ~1e5 facts for 5 dims of cardinalities: 32, 32, 49, 50, 1826
dc = as.data.cube(populate_star(N=1e5))
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(dc)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("data.cube: %.2f GB", as.numeric(object.size(dc$fact$data))/(1024^3))

# ~1e6 facts for 5 dims of cardinalities: 32, 32, 49, 50, 1826
dc = as.data.cube(populate_star(N=1e6))
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(dc)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("data.cube: %.2f GB", as.numeric(object.size(dc$fact$data))/(1024^3))

# ~1e6 facts for 5 dims of cardinalities: 32, 32, 49, 50, 3652
# twice bigger time dimension
dc = as.data.cube(populate_star(N=1e6, Y = c(2005L,2014L)))
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(dc)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("data.cube: %.2f GB", as.numeric(object.size(dc$fact$data))/(1024^3))
```
