---
vignette: >
  %\VignetteIndexEntry{Subset multidimensional data with cube}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Subset multidimensional data with cube
*Jan Gorecki, 2015-11-19*

R *cube* class defined in [data.cube](https://gitlab.com/jangorecki/data.cube) package.  

> Please note there is a new updated oop cube class called `data.cube`. See *Subset and aggregate multidimensional data with data.cube* vignette instead.

## characteristic of `[.cube`

### what's the same in `[.array`

- subsetting dimensions by keys
- handling *drop* argument
- returns same class object, so is *chainable*

### what's more than `[.array`

- can subset on hierarchy of a dimension
- scales with dimensions and their cardinality
- stores multiple measures
- high performance thanks to data.table

### what's different to `[.array`

- array can be subsetted by dimension keys but also by integer position
- in cube subset directly by integer position is not possible, but you can have integer surrogate keys which can refer to any custom integer positions

## start session

```{r run}
if(!"data.cube" %in% rownames(installed.packages())) install.packages(
    "data.cube", repos = paste0("https://", c("jangorecki.github.io/data.cube","cran.rstudio.com"))
)
library(data.table)
library(data.cube)
```

## tiny example

```{r tiny}
set.seed(1)
# array
ar = array(rnorm(8,10,5), 
           rep(2,3), 
           dimnames = list(color = c("green","red"), 
                           year = c("2014","2015"), 
                           country = c("UK","IN")))
# cube normalized to star schema
cb = as.cube(ar)

ar["green","2015",]
cb["green","2015",]

ar["green",c("2014","2015"),]
cb["green",c("2014","2015"),]

# tabular representation of array is just a formatting on the cube
format(cb["green",c("2014","2015"),], 
       dcast = TRUE, 
       formula = year ~ country)

ar[,"2015",c("UK","IN")]
cb[,"2015",c("UK","IN")]
format(cb[,"2015",c("UK","IN")], 
       dcast = TRUE, 
       formula = color ~ country)
```

## hierarchies example

```{r hierarchies}
# as.cube.list - investigate X to see structure
X = populate_star(N=1e5)
lapply(X, sapply, ncol)
lapply(X, sapply, nrow)
cb = as.cube(X)
print(cb)

# slice
cb["Mazda RX4"]
cb["Mazda RX4",,"BTC"]
# dice
cb[,, c("CNY","BTC"), c("GA","IA","AD")]

# check dimensions
cb$dims

# use dimensions hierarchy attributes for slice and dice, mix filters from various levels in hierarchy
cb["Mazda RX4",, .(curr_type = "crypto"),, .(time_year = 2014L, time_quarter_name = c("Q1","Q2"))]
# same as above but more verbose
cb[product = "Mazda RX4",
   customer = .(),
   currency = .(curr_type = "crypto"),
   geography = .(),
   time = .(time_year = 2014L, time_quarter_name = c("Q1","Q2"))]

# cube `[` operator returns another cube so queries can be chained
cb[,,, .(geog_region_name = "North Central")
   ][,,, .(geog_abb = c("IA","NV","MO")), .(time_year = 2014L)
     ]
```

## scalability

```{r scalability}
# ~1e5 facts for 5 dims of cardinalities: 32, 32, 49, 50, 1826
cb = as.cube(populate_star(N=1e5))
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(cb)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("cube: %.2f GB", as.numeric(object.size(cb$env$fact$sales))/(1024^3))

# ~1e6 facts for 5 dims of cardinalities: 32, 32, 49, 50, 1826
cb = as.cube(populate_star(N=1e6))
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(cb)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("cube: %.2f GB", as.numeric(object.size(cb$env$fact$sales))/(1024^3))

# ~1e6 facts for 5 dims of cardinalities: 32, 32, 49, 50, 3652
cb = as.cube(populate_star(N=1e6, Y = c(2005L,2014L))) # bigger time dimension
## estimated size of memory required to store an base R `array` for single numeric measure
sprintf("array: %.2f GB", (prod(dim(cb)) * 8)/(1024^3))
## fact table of *cube* object having multiple measures
sprintf("cube: %.2f GB", as.numeric(object.size(cb$env$fact$sales))/(1024^3))
```

## examples

Lots of examples can be found in tests: [tests/tests-sub-.cube.R](https://gitlab.com/jangorecki/data.cube/blob/master/tests/tests-sub-.cube.R).  
Feel free to PR your use case for future regression testing.  
